Normalization by Evaluation for Sized Dependent Types

### 2017-06-08 nbe/icfp17.tex
[ ICFP 17 ] Factored out lemma "fresh variable readback"

First, let us define some auxiliary judgements that relate a
well-formed syntactic object to a value, via read back.
They will constitute the logical relation for base types,
yet, are not general enough for function types.
\begin{alignat*}{8}
& \Gamma \der a && \doteq \tRsize\, \alpha
  && \quad\defas\quad
  && \forall \xi : \Gamma' \leq \Gamma.\ \
  && \mathrlap{ \RSize {\Gamma'} \alpha {a\xi} }
\\[-0.6ex]
& \Gamma \der T && \doteq \tRty\, A  : s
  && \quad\defas\quad
  && \forall \xi : \Gamma' \leq \Gamma.\ \
  && \exists V.\
  && \RTy{\Gamma'} A V
  && ~\mand~ \Gamma' \der T\xi = V : s
\\[-0.6ex]
& \Gamma \der t && \doteq \tR\, d  : T
  && \quad\defas\quad
  && \forall \xi : \Gamma' \leq \Gamma.\ \
  && \exists v.\
  && \R{\Gamma'} d v
  && ~\mand~ \Gamma' \der t\xi = v : T\xi
\\[-0.6ex]
& \Gamma \der t && \doteq \tRne\, n  : T
  && \quad\defas\quad
  && \forall \xi : \Gamma' \leq \Gamma.\ \
  && \exists m.\
  && \RNe{\Gamma'} n m
  && ~\mand~ \Gamma' \der t\xi = m : T\xi
\end{alignat*}
% \[
% \begin{array}{lll}
% \Gamma \der a \doteq \tRsize\, \alpha
%   & \defas &
%   \forall \xi : \Gamma' \leq \Gamma.\ \ \RSize {\Gamma'} \alpha {a\xi}
% \\
% \Gamma \der T \doteq \tRty\, A : s
%   & \defas &
%   \forall \xi : \Gamma' \leq \Gamma.\ \
%   \exists V.\
%   \RTy{\Gamma'} A V \mand
%   \Gamma' \der T\xi = V : s
% \\
% \Gamma \der t \doteq \tR\, d : T
%   & \defas &
%   \forall \xi : \Gamma' \leq \Gamma.\ \
%   \exists v.\
%   \R{\Gamma'} d v \mand
%   \Gamma' \der t\xi = v : T\xi
% \\
% \Gamma \der t \doteq \tRne\, n : T
%   & \defas &
%   \forall \xi : \Gamma' \leq \Gamma.\ \
%   \exists m.\
%   \RNe{\Gamma'} n m \mand
%   \Gamma' \der t\xi = m : T\xi
% \end{array}
% \]
% \[
%   \fbox{$\LRSize \Gamma a \alpha$} \defas
%   \forall \xi : \Gamma' \leq \Gamma.\ \
%   \RSize {\Gamma'} \alpha {a\xi}
%   % \exists a'.\
%   % \RSize {\Gamma'} \alpha a' \mand
%   % \Gamma' \der a\xi = a' : \Size
% \]
By definitions, these relations are closed under subsumption and
weakening, e.g., if\/ $\RD \Gamma t d T$ and $\Gamma \der T \leq T'$ then $\RD \Gamma t d T'$,
and if\/ $\xi : \Gamma' \leq \Gamma$ then $\RD{\Gamma'}{t\xi}d{T\xi}$.

\begin{lemma}[Closure properties for neutrals]
\label{lem:rclosne}
\bla
\begin{enumerate}
\item If\/ $\RNE \Gamma t n \piT U T$ and $\RD \Gamma u d U$ then $\RNE \Gamma {t\,u} {n\,d} {T[u]}$.
\item If\/ $\RNE \Gamma t n \piT \Size T$ and $\RSIZE \Gamma a \alpha$ then $\RNE \Gamma {t\,a} {n\,\alpha} {T[a]}$.
\item \label{it:closneall}
If\/ $\RNE \Gamma t n \forallT T$ and $\resurrect\Gamma \der a,b : \Size$ and $\alpha \in \SIZE$ then $\RNE \Gamma {t\,\ann a} {n\,\ann \alpha} {T[b]}$.
%% TOO WEAK:
% \item If\/ $\RNE \Gamma t n \forallT T$ and $\RSIZE {\resurrect\Gamma} a \alpha$ then $\RNE \Gamma {t\,\ann a} {n\,\ann \alpha} {T[a]}$.
\end{enumerate}
\end{lemma}
\LONGVERSION{
\begin{proof*}
For (\ref{it:closneall}), assume
$\xi : \Gamma' \leq \Gamma$.
We have to show that there exists a neutral normal form $m'$ such that
$\RNe {\Gamma'} {n\, \ann \alpha} m'$ and
$\Gamma' \der (t\, \ann a)\xi = m' : T[b]\xi$.
Sizes can always be read back, thus, let $\RSize {\Gamma'} \alpha {a_0}$.
By assumption $\RNE \Gamma t n {\forallT T}$, there is an $m$ with
$\RNe {\Gamma'} n m$ and
$\Gamma' \der t\xi = m : (\forallT T)\xi$.
The latter equation implies
$\Gamma' \der t\xi \ann {a\xi} = m \ann {a_0} : T(\xi,b\xi)$
by irrelevant size application to
$\resurrect{\Gamma'} \der a\xi, a_0, b \xi : \Size$,
thus we are done with $m' := m \ann{a_0}$.
\qed
\end{proof*}
} % END LONGVERSION

@@

\begin{enumerate}
\item
The premise is, after type conversion, $\Gamma \der t \doteq \tRne\, n : \piT U T$.
To demonstrate $\LR \Gamma t {T'} {\up {S'} n} {A'}$ we assume
$\xi : \Gamma' \leq \Gamma$ and $\LR{\Gamma'}{u}{U\xi} a A$ and show
$\LR{\Gamma'} {(t\xi\,u)} {T(\xi,u)} {\up{S[a]}(n\,\down A a)} {B[a]}$.
By induction hypothesis (\ref{it:out}) we have $\RD {\Gamma'} u {\down R a} {U\xi}$, and
together with the weakened premise $\Gamma' \der t\xi \doteq \tRne\, n : (\piT U T) \xi$ we get
$\Gamma' \der t\xi\, u \doteq \tRne\, (n\,\down R a) : T(\xi,u)$.
The goal follows by induction hypothesis (\ref{it:into}) for
% $B[a] \in s$.
$\LRTS{\Gamma'}{T(\xi,u)}{B[a]} s$.
% This follows by induction hypothesis (\ref{it:into}) for $B[a] \in \SET[\ell]$ if
% we can prove that for all $\xi' : \Gamma'' \leq \Gamma'$ there is a neutral normal form $m$
% such that $\RNe{\Gamma''} {(n\,\down A a)} m$ and $\Gamma'' \der (t\xi\,u)\xi' = m : T(\xi,u)\xi'$.
% Note that the last goal can be rewritten to $\Gamma'' \der t\xi\xi'\,(u\xi') = m : T(\xi\xi',u\xi')$.
% We know that $\RNe{\Gamma''} n {m_0}$ with $\Gamma'' \der t\xi\xi' = m_0 : T'\xi\xi'$.
% Thus, it is sufficient to show that there is some normal form $v$ such that
% $\R{\Gamma''} {\down A a} v$ and $\Gamma'' \der u\xi' = v : U\xi\xi'$.
% However, this follows by induction hypothesis (\ref{it:out}) on
% assumption $\LR{\Gamma'}{u}{U\xi} a A$ with $\xi'$.
\end{enumerate}

@@

\begin{enumerate}
\item We assume $\xi : \Gamma' \leq \Gamma$ and
$\resurrect{\Gamma'} \der a, b : \Size$ and $\alpha,\beta \in \SIZE$ with
$\LRSize {\resurrect{\Gamma'}} b \beta$ and show
$\LR {\Gamma'} {t\xi\ann a} {T(\xi,b)} {(\up {S'} n) \ann \alpha} {A[\beta]}$.
By the evaluation rules for reflection, it is sufficient to show
% It is sufficient to show
$\LR {\Gamma'} {t\xi\ann a} {T(\xi,b)} {\up {S[\alpha]} (n \ann \alpha)} {A[\beta]}$.
This follows by induction hypothesis (\ref{it:into}) if
$\LRTS {\Gamma'} {T(\xi,b)} {A[\beta]} s$ and $A[\beta] \shape S[\alpha]$ and
$\RNE {\Gamma'} {t\xi\ann a} {n \ann \alpha} {T(\xi,b)}$.
The first two of these subgoals are immediate.
The third follows by Lemma~\ref{lem:rclosne} (\ref{it:closneall}) from the
weakened assumption
 $\RNE {\Gamma'} {t\xi} n {T'\xi}$.
% For the third, assume
% $\xi' : \Gamma'' \leq \Gamma'$.
% We have to show that there exists a neutral normal form $m'$ such that
% $\RNe {\Gamma''} {n \ann \alpha} m'$ and
% $\Gamma'' \der (t\xi \ann a)\xi' = m' : (T(\xi,b))\xi'$.
% Sizes can always be read back, thus, let $\RSize {\Gamma''} \alpha {a_0}$.
% By assmption $\RNE \Gamma t n {T'}$, there is an $m$ with
% $\RNe {\Gamma''} n m$ and
% $\Gamma'' \der t\xi\xi' = m : T'\xi\xi'$.
% The latter equation implies
% $\Gamma'' \der t\xi\xi' \ann {a\xi'} = m \ann {a_0} : T(\xi\xi',b\xi')$
% by irrelevant size application to
% $\resurrect{\Gamma''} \der a\xi', a_0, b \xi' : \Size$,
% thus we are done by choosing $m' : = m \ann{a_0}$.
%
\qed
\end{enumerate}
%
\end{caselist}
\end{proof*}
} % END LONGVERSION

\LONGVERSION{
\begin{corollary}[One-to-one]
\label{cor:oneone}
\bla
\begin{enumerate}
\item
If\/ $\LRSize \Gamma a \alpha$ and $\LRSize \Gamma {a'} \alpha$ then $a = a'$.
\item
If\/ $\LRT \Gamma T A \ell$ and $\LRT \Gamma {T'} A {\ell'}$ then $\Gamma \der T = T'$.
\end{enumerate}
\end{corollary}
} % END LONGVERSION
\begin{corollary}[Fresh variable]
  \label{cor:fresh}
  If\/ $\der \cext \Gamma T$ and %$A \in \SET[\ell]$
  $\RG \Gamma T A$
  then $\LR {\cext \Gamma T} {\ind 0} {(T \slift)} {(\up A \var\Gamma)} A$.
\end{corollary}
\begin{proof}
  Assume $\xi : \Gamma' \leq \cext \Gamma T$.
  Note that for the length
  $i = |\Gamma'| - |\cext \Gamma T|$ of the context extension
  we have $\ind i = \ind 0 \xi$, thus
  $\Gamma' \der \ind 0 \xi = \ind i : T \slift \xi$.
  Since $\RNe{\Gamma'} {\var\Gamma} {\ind i}$,
  the goal follows now from Thm.~\ref{thm:into} part (\ref{it:into}).
\end{proof}

>>>>

First, let us define some auxiliary judgements that relate a
well-formed syntactic object to a value, via read back.
They will constitute the logical relation for base types,
yet, are not general enough for function types.
\begin{alignat*}{8}
& \Gamma \der a && \doteq \tRsize\, \alpha
  && \quad\defas\quad
  && \forall \xi : \Gamma' \leq \Gamma.\ \
  && \mathrlap{ \RSize {\Gamma'} \alpha {a\xi} }
\\[-0.6ex]
& \Gamma \der T && \doteq \tRty\, A  : s
  && \quad\defas\quad
  && \forall \xi : \Gamma' \leq \Gamma.\ \
  && \exists V.\
  && \RTy{\Gamma'} A V
  && ~\mand~ \Gamma' \der T\xi = V : s
\\[-0.6ex]
& \Gamma \der t && \doteq \tR\, d  : T
  && \quad\defas\quad
  && \forall \xi : \Gamma' \leq \Gamma.\ \
  && \exists v.\
  && \R{\Gamma'} d v
  && ~\mand~ \Gamma' \der t\xi = v : T\xi
\\[-0.6ex]
& \Gamma \der t && \doteq \tRne\, n  : T
  && \quad\defas\quad
  && \forall \xi : \Gamma' \leq \Gamma.\ \
  && \exists m.\
  && \RNe{\Gamma'} n m
  && ~\mand~ \Gamma' \der t\xi = m : T\xi
\end{alignat*}
% \[
% \begin{array}{lll}
% \Gamma \der a \doteq \tRsize\, \alpha
%   & \defas &
%   \forall \xi : \Gamma' \leq \Gamma.\ \ \RSize {\Gamma'} \alpha {a\xi}
% \\
% \Gamma \der T \doteq \tRty\, A : s
%   & \defas &
%   \forall \xi : \Gamma' \leq \Gamma.\ \
%   \exists V.\
%   \RTy{\Gamma'} A V \mand
%   \Gamma' \der T\xi = V : s
% \\
% \Gamma \der t \doteq \tR\, d : T
%   & \defas &
%   \forall \xi : \Gamma' \leq \Gamma.\ \
%   \exists v.\
%   \R{\Gamma'} d v \mand
%   \Gamma' \der t\xi = v : T\xi
% \\
% \Gamma \der t \doteq \tRne\, n : T
%   & \defas &
%   \forall \xi : \Gamma' \leq \Gamma.\ \
%   \exists m.\
%   \RNe{\Gamma'} n m \mand
%   \Gamma' \der t\xi = m : T\xi
% \end{array}
% \]
% \[
%   \fbox{$\LRSize \Gamma a \alpha$} \defas
%   \forall \xi : \Gamma' \leq \Gamma.\ \
%   \RSize {\Gamma'} \alpha {a\xi}
%   % \exists a'.\
%   % \RSize {\Gamma'} \alpha a' \mand
%   % \Gamma' \der a\xi = a' : \Size
% \]
By definitions, these relations are closed under subsumption and
weakening, e.g., if\/ $\RD \Gamma t d T$ and $\Gamma \der T \leq T'$ then $\RD \Gamma t d T'$,
and if\/ $\xi : \Gamma' \leq \Gamma$ then $\RD{\Gamma'}{t\xi}d{T\xi}$.
\begin{lemma}[Fresh variable readback]
  \label{lem:freshread}
  If\/ $\Gamma \der U$ then $\RNE{\cext \Gamma U} {\ind 0} {\var \Gamma} {U \slift}$.
\end{lemma}
\begin{proof}
  Assume $\xi : \Gamma' \leq \cext \Gamma U$.
  Note that $\RNe{\Gamma'} {\var\Gamma} {\ind i}$ where
  $i = |\Gamma'| - |\cext \Gamma U|$ is the length of the context extension.
  Since $\ind 0 \xi = \ind i$, we conclude
  $\Gamma' \der \ind 0 \xi = \ind i : U \slift \xi$
  by weakening of the judgement $\cext \Gamma U \der \ind 0 = \ind 0 : U \slift$.
\end{proof}
\begin{lemma}[Closure properties for neutrals]
\label{lem:rclosne}
\bla
\begin{enumerate}
\item If\/ $\RNE \Gamma t n \piT U T$ and $\RD \Gamma u d U$ then $\RNE \Gamma {t\,u} {n\,d} {T[u]}$.
\item If\/ $\RNE \Gamma t n \piT \Size T$ and $\RSIZE \Gamma a \alpha$ then $\RNE \Gamma {t\,a} {n\,\alpha} {T[a]}$.
\item \label{it:closneall}
If\/ $\RNE \Gamma t n \forallT T$ and $\resurrect\Gamma \der a,b : \Size$ and $\alpha \in \SIZE$ then $\RNE \Gamma {t\,\ann a} {n\,\ann \alpha} {T[b]}$.
%% TOO WEAK:
% \item If\/ $\RNE \Gamma t n \forallT T$ and $\RSIZE {\resurrect\Gamma} a \alpha$ then $\RNE \Gamma {t\,\ann a} {n\,\ann \alpha} {T[a]}$.
\end{enumerate}
\end{lemma}
\LONGVERSION{
\begin{proof*}
For (\ref{it:closneall}), assume
$\xi : \Gamma' \leq \Gamma$.
We have to show that there exists a neutral normal form $m'$ such that
$\RNe {\Gamma'} {n\, \ann \alpha} m'$ and
$\Gamma' \der (t\, \ann a)\xi = m' : T[b]\xi$.
Sizes can always be read back, thus, let $\RSize {\Gamma'} \alpha {a_0}$,
which guarantees $\resurrect{\Gamma'} \der a_0 : \Size$.
By assumption $\RNE \Gamma t n {\forallT T}$, there is an $m$ with
$\RNe {\Gamma'} n m$ and
$\Gamma' \der t\xi = m : (\forallT T)\xi$.
The latter equation implies
$\Gamma' \der t\xi \ann {a\xi} = m \ann {a_0} : T(\xi,b\xi)$
by irrelevant size application to
$\resurrect{\Gamma'} \der a\xi, a_0, b \xi : \Size$,
thus we are done with $m' := m \ann{a_0}$.
\qed
\end{proof*}
} % END LONGVERSION

@@

\begin{enumerate}
\item % into
The premise is, after type conversion, $\Gamma \der t \doteq \tRne\, n : \piT U T$.
To demonstrate $\LR \Gamma t {T'} {\up {S'} n} {A'}$ we assume
$\xi : \Gamma' \leq \Gamma$ and $\LR{\Gamma'}{u}{U\xi} a A$ and show
$\LR{\Gamma'} {(t\xi\,u)} {T(\xi,u)} {\up{S[a]}(n\,\down A a)} {B[a]}$.
By induction hypothesis (\ref{it:out}) we have $\RD {\Gamma'} u {\down R a} {U\xi}$, and
together with the weakened premise $\Gamma' \der t\xi \doteq \tRne\, n : (\piT U T) \xi$ we get
$\Gamma' \der t\xi\, u \doteq \tRne\, (n\,\down R a) : T(\xi,u)$.
The goal follows by induction hypothesis (\ref{it:into}) for
% $B[a] \in s$.
$\LRTS{\Gamma'}{T(\xi,u)}{B[a]} s$.
% This follows by induction hypothesis (\ref{it:into}) for $B[a] \in \SET[\ell]$ if
% we can prove that for all $\xi' : \Gamma'' \leq \Gamma'$ there is a neutral normal form $m$
% such that $\RNe{\Gamma''} {(n\,\down A a)} m$ and $\Gamma'' \der (t\xi\,u)\xi' = m : T(\xi,u)\xi'$.
% Note that the last goal can be rewritten to $\Gamma'' \der t\xi\xi'\,(u\xi') = m : T(\xi\xi',u\xi')$.
% We know that $\RNe{\Gamma''} n {m_0}$ with $\Gamma'' \der t\xi\xi' = m_0 : T'\xi\xi'$.
% Thus, it is sufficient to show that there is some normal form $v$ such that
% $\R{\Gamma''} {\down A a} v$ and $\Gamma'' \der u\xi' = v : U\xi\xi'$.
% However, this follows by induction hypothesis (\ref{it:out}) on
% assumption $\LR{\Gamma'}{u}{U\xi} a A$ with $\xi'$.
\end{enumerate}

@@

\begin{enumerate}
\item % into
We assume $\xi : \Gamma' \leq \Gamma$ and
$\resurrect{\Gamma'} \der a, b : \Size$ and $\alpha,\beta \in \SIZE$ with
$\LRSize {\resurrect{\Gamma'}} b \beta$ and show
$\LR {\Gamma'} {t\xi\ann a} {T(\xi,b)} {(\up {S'} n) \ann \alpha} {A[\beta]}$.
By the evaluation rules for reflection, it is sufficient to show
% It is sufficient to show
$\LR {\Gamma'} {t\xi\ann a} {T(\xi,b)} {\up {S[\alpha]} (n \ann \alpha)} {A[\beta]}$.
This follows by induction hypothesis (\ref{it:into}) if
$\LRTS {\Gamma'} {T(\xi,b)} {A[\beta]} s$ and $A[\beta] \shape S[\alpha]$ and
$\RNE {\Gamma'} {t\xi\ann a} {n \ann \alpha} {T(\xi,b)}$.
The first two of these subgoals are immediate.
The third follows by Lemma~\ref{lem:rclosne} (\ref{it:closneall}) from the
weakened assumption
 $\RNE {\Gamma'} {t\xi} n {T'\xi}$.
% For the third, assume
% $\xi' : \Gamma'' \leq \Gamma'$.
% We have to show that there exists a neutral normal form $m'$ such that
% $\RNe {\Gamma''} {n \ann \alpha} m'$ and
% $\Gamma'' \der (t\xi \ann a)\xi' = m' : (T(\xi,b))\xi'$.
% Sizes can always be read back, thus, let $\RSize {\Gamma''} \alpha {a_0}$.
% By assmption $\RNE \Gamma t n {T'}$, there is an $m$ with
% $\RNe {\Gamma''} n m$ and
% $\Gamma'' \der t\xi\xi' = m : T'\xi\xi'$.
% The latter equation implies
% $\Gamma'' \der t\xi\xi' \ann {a\xi'} = m \ann {a_0} : T(\xi\xi',b\xi')$
% by irrelevant size application to
% $\resurrect{\Gamma''} \der a\xi', a_0, b \xi' : \Size$,
% thus we are done by choosing $m' : = m \ann{a_0}$.
%
\qed
\end{enumerate}
%
\end{caselist}
\end{proof*}
} % END LONGVERSION
\begin{corollary}[Fresh variable injection]
  \label{cor:fresh}
  If\/ $\der \cext \Gamma T$ and %$A \in \SET[\ell]$
  $\RG \Gamma T A$
  then $\LR {\cext \Gamma T} {\ind 0} {(T \slift)} {(\up A \var\Gamma)} A$.
\end{corollary}
\begin{proof}
  Since $\RNE {\cext \Gamma T} {\ind 0} {\var \Gamma} {T \slift}$
  by Lemma~\ref{lem:freshread},
  the goal follows from Thm.~\ref{thm:into} part (\ref{it:into}).
  % Assume $\xi : \Gamma' \leq \cext \Gamma T$.
  % Note that for the length
  % $i = |\Gamma'| - |\cext \Gamma T|$ of the context extension
  % we have $\ind i = \ind 0 \xi$, thus
  % $\Gamma' \der \ind 0 \xi = \ind i : T \slift \xi$.
  % Since $\RNe{\Gamma'} {\var\Gamma} {\ind i}$,
  % the goal follows now from Thm.~\ref{thm:into} part (\ref{it:into}).
\end{proof}

\LONGVERSION{
\begin{corollary}[One-to-one]
\label{cor:oneone}
\bla
\begin{enumerate}
\item
If\/ $\LRSize \Gamma a \alpha$ and $\LRSize \Gamma {a'} \alpha$ then $a = a'$.
\item
If\/ $\LRT \Gamma T A \ell$ and $\LRT \Gamma {T'} A {\ell'}$ then $\Gamma \der T = T'$.
\end{enumerate}
\end{corollary}
} % END LONGVERSION
